Make any stupid plot you want

### 2021-05-05 draft/sections/intro.tex
conjectures, no homomorphisms

\subsection{Structure Visualizations Preserve}
\label{sec:intro:structure}

\begin{figure}[H]
        \includegraphics[width=1\textwidth]{figures/intro/equivariant.png}
        \caption{The precipitation in blue is scaled by a factor of 10, yielding the precipitation in orange in the table (a). This scaling factor is represented by the green arrows in (a). The \textit{equivariant} action is the scaling in the timeseries (b), which is represented by the green fill. Because the scalings are equivariant, the orange precipitation timeseries is 10x the blue precipitation timeseries. The red line is an example of \textit{equivariance} not being preserved, as it is not 10x the the y value of the blue precipitation timeseries.}
      \label{fig:intro:equivariance}
\end{figure}
The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}. Bertin proposes that there are categories of visual encodings-such as position, shape, color, and texture-that preserve the properties of the measurement type, quantitative or qualitative, of the encoded data. For example, in \autoref{fig:intro:equivariance}, the blue precipitation data in the table (a) is scaled by a factor of 10. This scaling action is represented by the green arrows that translate the precipitation into the scaled values in orange. For this visualization to be equivariant, this same scaling factor must be present in the timeseries representation of the data (b). The precipitation in orange is an \textit{equivariant} scaling of the precipitation in blue, meaning that the y-values of the orange line is 10x the y-values of the blue precipitation timeseries. The green fill in the timeseries (b) is the scaling equivalent to the arrows in the table (a). In contrast, the red line is not an \textit{equivariant} representation to the 10x scaling because the red y-values are not 10x the blue y values. The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}. The algebraic model of visualization proposed by Kindlmann and Scheidegger uses equivariance to refer generally to invertible binary transformations\cite{kindlmannAlgebraicProcessVisualization2014}, which are mathematical groups \cite{shadrachIntroductionGroups2017}. Our model defines \textit{equivariance} in terms of monoid actions, which are a more restrictive set than all binary operations and more general than groups. As with the algebraic model, our model also defines structure preservation as commutative mappings from data space to representation space to graphic space, but our model uses topology to explicitely include continuity.

>>>>

\subsection{Structure}
\label{sec:intro:structure}

\begin{figure}[H]
        \includegraphics[width=1\textwidth]{figures/intro/equivariant.png}
        \caption{The table (a) lists precipitation in millimeters in blue. These values are converted to  inches, in green, by dividing the millimeter values by 25.4. This conversion is a scaling action, represented by the orange arrows in (a). The \textit{equivariant} action is the scaling in the timeseries (b), which is represented by the orange fill between the green and blue lines. In (c), the scaling is illustrated in the different millimeter and inches y axis labeling of data that appears identical since the distances were scaled by a constant factor that is now incorporated in the labeling.}
      \label{fig:intro:equivariance}
\end{figure}
The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}. Bertin proposes that there are categories of visual encodings-such as position, shape, color, and texture-that preserve the properties of the measurement type, quantitative or qualitative, of the encoded data. For example, in \autoref{fig:intro:equivariance}, the blue precipitation data in millimeters in the table (a) is converted to inches. This scaling action is represented by the green arrows that translate the precipitation into the scaled values in green. For this visualization to be equivariant, this same scaling factor must be present in the timeseries representation of the data (b). The precipitation in green is an \textit{equivariant} scaling of the precipitation in blue, meaning that the y-values of the green line is $\frac{1}{25.4}$ that of the y-values of the blue precipitation timeseries. The orange fill in the timeseries (b) is the scaling equivalent to the arrows in the table (a). By definition, the distances in the millimeter data were scaled equally \cite{stothersSimilarityGroup} when converted to inches, which means the shapes of the graphs are equivalent when plotted against y-axes that preserve relative distance (c). This visualization is also equivariant to the table (a), but this is indicated through labeling rather than the line plots. The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}. The algebraic model of visualization proposed by Kindlmann and Scheidegger uses equivariance to refer generally to invertible binary transformations\cite{kindlmannAlgebraicProcessVisualization2014}, which are mathematical groups \cite{shadrachIntroductionGroups2017}. Our model defines \textit{equivariance} in terms of monoid actions, which are a more restrictive set than all binary operations and more general than groups. As with the algebraic model, our model also defines structure preservation as commutative mappings from data space to representation space to graphic space, but our model uses topology to explicitly include continuity.

### 2021-05-05 draft/sections/math.tex
conjectures, no homomorphisms

For example, the records in the 2D fiber (a) in \autoref{fig:math:data:fiber} are a pair of \textit{times} and \textit{precipitation} measurements taken at those times. Time is a positive number of type \texttt{datetime} which can be resolved to $\ftotal_{\texttt{datetime}}= \reals$. Precipitation values are real positive numbers $\ftotal_{\texttt{float}} = \realsp$. The fiber is 
\begin{equation*}
    \dfiber =  \reals \times \realsp 
\end{equation*} 
where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals)$ and is the set of values $\fttype=\reals$. In the 3D fiber (b) in \autoref{fig:math:data:fiber}, time is replaced with location. This location variable is of type \texttt{point} and has two components \textit{latitude} and \textit{longitude} $\{(lat,lon) \in \reals^{2} \mid  -90\leq lat \leq 90,\, 0 \leq lon \leq 360\}$. The fiber for this dataset is
\begin{equation*}
    \dfiber = \reals \times [0, 360] \times [-90, 90]
\end{equation*} 
and can either be thought of as having two components  
\begin{equation}
    \{(\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals), ((\fname = location,\, \ftype = \texttt{point},\, \fttype=\reals^{2})\}
\end{equation}
 or three
 \begin{equation}
    \{(\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals), (\fname=latitude,\, \ftype=\texttt{float}, \, \fttype=\reals), (\fname=longitude,\, \ftype=\texttt{float},\,\fttype=\reals)\}
 \end{equation}

\subsubsection{Measurement Scales: Monoid Actions}
\label{sec:math:data:monoid}
Implementing expressive visual encodings requires formally describing the structure on the components of the fiber, which we define by the actions of a monoid on the component. In doing so, we specify the properties of the component that must be preserved in a graphic representation. A monoid \cite{Monoid2021} $\monoid$ is a set with a binary operation $\ast:\monoid \times \monoid\rightarrow \monoid$ that satisfies the axioms:
\begin{align*}
    \textbf{associativity}\; & \text{for all} a, b, c \in \monoid \, (a \bullet b) \bullet c = a \bullet (b \bullet c)\\
    \textbf{identity}\; & \text{for all } a\in \monoid,\,  e\bullet a = a 
\end{align*} 
As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
\end{align*}
The identity and associativity properties of the action denote that the action is a monoid homomorphism, which means that the group operation is preserved on both sides of the action\cite{weissteinGroupHomomorphism}.

\begin{figure}[H]
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/monoid_total.png}
        \caption{}
        \label{fig:math:data:monoid:total}
    \end{subfigure}
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=1\textwidth]{figures/math/monoid_partial.png}
        \caption{}
        \label{fig:math:data:monoid:partial}
    \end{subfigure}
    \caption{The action $\bullet$ in \autoref{fig:monoid:total} is the arrows from the partial order diagram of weather states on the left to the diagram of weather states on the right. Since the action maps the weather states to themselves, the ordering defined by the monoid $\ast$ is preserved on both sides of the action. The action in \autoref{fig:monoid:partial} is monoid homomorphism because the ordering of the weather states is the same as the ordering of the elements they are mapped to. Given $sun \geq clouds\geq rain\, lightining$ on the right, the action  $sun\, clouds \rightarrow clouds$, and $rain\, lightining \rightarrow rain$ is structure preserving because on the left $cloud \geq rain$ so the relative ordering of elements is the same as the elements they are mapped to.}
    \label{fig:math:data:monoid}
\end{figure}

One example of monoids are partial orderings on a set, such as seen in \autoref{fig:math:data:monoid}. Each hasse diagram of the set of weather states describes an ordering on the set; the arrow goes from the lesser value to the greater one. For example, $storm \leq rain$. In \autoref{fig:monoid:total}, the action $\bullet$ maps the elements of a set of weather states into itself by mapping them into other elements of the weather states. The action in  \autoref{fig:math:data:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action and it is therefore homomorphic. The action $\bullet$ in \autoref{fig:math:data:monoid:partial} is a monotone map\cite{fongInvitationAppliedCategory2019}
\begin{equation*}
    if\, a \leq b \; then \; \bullet(a) \leq \bullet(b) \mid a,b \in \dfiber_{i}
\end{equation*}
where the structure the action preserves is the relative, rather than exact, ordering. Since groups are monoids with invertible operations, this definition of structure is also broad enough to include the Steven's measurment scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}. Monoids are also commonly found in functional programming since the core property of monoids is composability \cite{yorgeyMonoidsThemeVariations}.

@@

\subsubsection{Assembly \vmark}
Given the continuities described in \ref{fig:math:graphic:retraction:map}, 
we illustrate a minimal Q that will generate the most minimal visualizations associated with those continuities: non-overlapping scatter points, a non-infinitely thin line, and an image. 
\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/math/scatter_with_s.png}
    \caption{The data is discrete points (precipitation, time). Via \vchannel\ these are converted to (xpos, ypos) and pulled over discrete \gbase. These values are then used to parameterize \gsection\ which returns a color based on the parameters (xpos,ypos) and position $\alpha, \beta$ on $\gbase_{\dbasepoint}$ that \gsection\ is evaluated on. 
    }
    \label{fig:math:artist:q:scatter}
\end{figure}
The scatter plot in \autoref{fig:math:artist:q:scatter} has a constant size and color $\gsection_{RGB} = (0,0,0)$ that are defined as part of the point assembly function.

@@

\begin{figure}[H]
    \includegraphics[width=\textwidth]{figures/math/line_with_s.png}
    \caption{The line fiber $(time,\, temp)$ is thickened with the derivative $(time^{\prime},\, precipitation^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line. This is because the line needs to be pushed perpendicular to the tangent of (xpos, ypos). The data is converted to visual characteristics (xpos, ypos). The $\alpha$ coordinates on \gbase\ specifies the position of the line, the $\beta$ coordinate specifies thickness.}
    \label{fig:math:artist:q:line}
\end{figure}
In contrast, the line plot in \autoref{fig:math:artist:q:line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.

>>>>

For example, the records in the 2D fiber (a) in \autoref{fig:math:data:fiber} are a pair of \textit{times} and \textit{precipitation} measurements taken at those times. Time is a positive number of type \texttt{datetime} which can be resolved to $\ftotal_{\texttt{datetime}}= \reals$. Precipitation values are real positive numbers $\ftotal_{\texttt{float}} = \realsp$. The fiber is 
\begin{equation*}
    \dfiber =  \reals \times \realsp 
\end{equation*} 
where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals)$ and is the set of values $\fttype=\reals$. In the 3D fiber (b) in \autoref{fig:math:data:fiber}, time is replaced with location. This location variable is of type \texttt{point} and has two components \textit{latitude} and \textit{longitude} $\{(lat,lon) \in \reals^{2} \mid  -90\leq lat \leq 90,\, 0 \leq lon \leq 360\}$. The fiber for this dataset is
\begin{equation*}
    \dfiber = \reals \times \left[0, 360\right] \times \left[-90, 90\right]
\end{equation*} 
with components \((\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals)\), \((\fname=latitude,\, \ftype=\texttt{float}, \, \fttype=\left[0, 360\right])\), and \((\fname=longitude,\, \ftype=\texttt{float},\,\fttype= \left[-90, 90\right]\). By adapting Spivak's framework, our model has a consistent way to describe the components of the data, no matter their complexity.

\subsubsection{Measurement Scales: Monoid Actions}
\label{sec:math:data:monoid}
Implementing expressive visual encodings requires formally describing the structure on the components of the fiber. We conjecture that the structure can be formally defined as the actions of a monoid on the component. We can also discuss complex structures since a core property of monoids is composability \cite{yorgeyMonoidsThemeVariations}. In doing so, we can specify the properties of the component that must be preserved in a graphic representation.

A monoid \cite{Monoid2021} $\monoid$ is a set with a binary operation $\ast:\monoid \times \monoid\rightarrow \monoid$ that satisfies the axioms:
\begin{align*}
    \textbf{associativity}\; & \text{for all} a, b, c \in \monoid \, (a \bullet b) \bullet c = a \bullet (b \bullet c)\\
    \textbf{identity}\; & \text{for all } a\in \monoid,\,  e\bullet a = a 
\end{align*} 
As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
\end{align*}
We conjecture that given these properties, if there is a partial ordering on both sides of the action then the action is equivariant and monotonic.

\begin{figure}[H]
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/monoid_total.png}
        \caption{}
        \label{fig:math:data:monoid:total}
    \end{subfigure}
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=1\textwidth]{figures/math/monoid_partial.png}
        \caption{}
        \label{fig:math:data:monoid:partial}
    \end{subfigure}
    \caption{The action $\bullet$ in \autoref{fig:math:data:monoid:total} is the arrows from the partial order diagram of weather states on the left to the diagram of weather states on the right. The action maps the weather states to themselves, thereby preserving the ordering defined by the monoid $\ast$ on both sides of the action. We conjecture that the action in \autoref{fig:math:data:monoid:partial} is monotonic when the partial ordering of the weather states is the same as the partial order of the elements they are mapped to. Given $sun \geq clouds\geq rain,\, lightining$ on the right, we conjecture that the action  $sun,\, clouds \rightarrow clouds$, and $rain\, lightining \rightarrow rain$ is structure preserving such that the relative ordering of elements is the same as the elements they are mapped to.}
    \label{fig:math:data:monoid}
\end{figure}

One example of monoids are partial orderings on a set, such as seen in \autoref{fig:math:data:monoid}. Each hasse diagram of the set of weather states describes an ordering on the set; the arrow goes from the lesser value to the greater one. For example, $storm \leq rain$. In \autoref{fig:math:data:monoid}, the action $\bullet$ maps the elements of a set of weather states into itself by mapping them into other elements of the weather states. The action in  \autoref{fig:math:data:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action. The action $\bullet$ in \autoref{fig:math:data:monoid:partial} is a monotone map\cite{fongInvitationAppliedCategory2019}
\begin{equation*}
    if\, a \leq b \; then \; \bullet(a) \leq \bullet(b) \mid a,b \in \dfiber_{i}
\end{equation*}
where the structure the action preserves is the relative, rather than exact, ordering. Since groups are monoids with invertible operations, this definition of structure is broad enough to include the Steven's measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.

@@

\subsubsection{Assembly \vmark}
Given the continuities described in \ref{fig:math:graphic:retraction:map}, 
we illustrate a minimal Q that will generate the most minimal visualizations associated with those continuities: non-overlapping scatter points, a non-infinitely thin line, and an image. 
\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/math/scatter.png}
    \caption{The data is discrete points (precipitation, time). Via \vchannel\ these are converted to (xpos, ypos) and pulled over discrete \gbase. These values are then used to parameterize \gsection\ which returns a color based on the parameters (xpos,ypos) and position $\alpha, \beta$ on $\gbase_{\dbasepoint}$ that \gsection\ is evaluated on. 
    }
    \label{fig:math:artist:q:scatter}
\end{figure}
The scatter plot in \autoref{fig:math:artist:q:scatter} has a constant size and color $\gsection_{RGB} = (0,0,0)$ that are defined as part of the point assembly function.

@@

\begin{figure}[H]
    \includegraphics[width=\textwidth]{figures/math/line.png}
    \caption{The line fiber $(time,\, temp)$ is thickened with the derivative $(time^{\prime},\, precipitation^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line. This is because the line needs to be pushed perpendicular to the tangent of (xpos, ypos). The data is converted to visual characteristics (xpos, ypos). The $\alpha$ coordinates on \gbase\ specifies the position of the line, the $\beta$ coordinate specifies thickness.}
    \label{fig:math:artist:q:line}
\end{figure}
In contrast, the line plot in \autoref{fig:math:artist:q:line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
